Evaluating the Performance and Energy Efficiency of COSMO-ART, a Coupled Numerical Weather Forecast and Chemical Transport Model

### 2014-04-18 paper_enahpc14/Introduction.tex
adds very few changes; adds Cristiano Malossi as a co-author

Some  related work  will  be discussed  in  Sec.~\ref{sec:1}. Then  in
Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
COSMO-ART and specify its  technical setup related to the investigated
performance and energy evaluation methods.  Sec.~\ref{sec:3} describes
different  power  measurement  equipments  used  for  conducting  this
benchmarking  study and presents  energy-aware techniques  employed to
determine  the   energy  footprint  of   the  considered  application.
Sec.~\ref{sec:4} gathers benchmark  results from the Exa2Green project
and  highlights areas  where improvements  will be  necessary  for the
subsequent baseline refactoring.   Finally, we conclude with practical
guidelines  to  identifying   energy  saving  opportunities  and  give
directions  for  further  developments   of  the  modeling  system  in
Sec.~\ref{concl}.

>>>>

Some  related work  will  be discussed  in  Sec.~\ref{sec:1}. Then  in
Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
COSMO-ART and specify its  technical setup related to the investigated
performance and energy evaluation methods.  Sec.~\ref{sec:3} describes
different  power  measurement  equipments  used  for  conducting  this
benchmarking  study and presents  energy-aware techniques  employed to
determine  the   energy  footprint  of   the  considered  application.
Sec.~\ref{sec:4} gathers benchmark  results from the Exa2Green project
and  highlights areas  where improvements  will be  necessary  for the
subsequent baseline refactoring.   Finally, we conclude with practical
guidelines  to identifying  energy  saving opportunities  and give  an
outlook to future research in Sec.~\ref{concl}.

### 2014-04-18 paper_enahpc14/Sec4:Evaluation.tex
adds very few changes; adds Cristiano Malossi as a co-author

\subsection{Environment setup}
\label{subsec:4.2}
While hardware  platforms mature  and are replaced,  we have  chosen a
state-of-the-art  Intel's  third-generation   Core  (aka  Ivy  Bridge)
processing platform (called ``Monch'')  for our power measurements, as
it  is slated  to stay  in service  without hardware  upgrade  for the
duration  of  the Exa2Green  project.   In  principle  at least,  this
architecture could be recreated or found in an identical configuration
beyond the lifetime of the project.  Given that the baseline benchmark
can be reproduced  within an expected variance, and  that the baseline
run configuration  can be used in  all future versions of  the code, a
fair comparison  will be made  between the baseline and  the milestone
versions   of    COSMO-ART.    In   this    study,   a   complementary
energy-to-solution  benchmarking  comparison  is  carried out  on  the
``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
processors, conventional  in HPC  systems and known  to be  more power
consuming.

\subsubsection{Monch (CSCS - ETH Zurich)}
Monch, which is installed  at the Swiss National Supercomputing Center
(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
Ivy Bridge-EP based cluster, utilised from people that are part of the
Swiss    Platform   for    Advanced   Scientific    Computing   (PASC,
\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
Each standard  compute node comprises two Intel  Ivy Bridge-EP E5-2660
v2 ten-core processors operating at 2.2 GHz, themselves connected by a
high speed  InfiniBand network, based  on Mellanox SX6036  managed FDR
switches, with a  56 Gb/s speed.  Each core has  32 KB instruction and
32 KB data L1  caches and 256 KB of L2 cache. All  the 10 core share a
25 MB L3 cache and the platform has 32GB of DDR3 1600 MHz RAM. For our
energy-to-solution benchmark,  a full rack of Monch  constituted of 52
standard  compute   nodes  (monchc[029-080])  was   considered.

\subsubsection{Pilatus (CSCS - ETH Zurich)}
Pilatus,  which  is installed  at  the  Swiss National  Supercomputing
Center (CSCS)  of ETH Zurich,  is a dual-socket Intel  Sandy Bridge-EP
based cluster  used as Piz  Daint pre-post processing cluster.   It is
composed of 42 compute nodes  and has 2 high-speed interconnects based
on FDR:  the first is dedicated to  the MPI traffic and  the second to
the storage high speed traffic.  The 2 login nodes and the 42 computes
nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
scalable compute modules.  Each  module contains 4 compute nodes based
on two Intel Xeon E5-2670  eight-core processors operating at 2.6 GHz,
themselves  connected by  a high  speed InfiniBand  network,  based on
Mellanox SX6036 managed FDR switches,  with a 56 Gb/s speed. Each core
has 32 KB instruction and 32 KB data L1 caches and 256 KB of L2 cache.
All the  8 core share a  20 MB L3 cache  and the platform  has 64GB of
DDR3 1600 MHz  RAM. For our energy-to-solution benchmark,  a full rack
of Pilatus  constituted of 42 standard  compute nodes (pilatus[03-44])
was considered.

@@

\subsubsection{Sofware environment}
INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
distributed  memory  parallel  computers  using  the  Message  Passing
Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
both many-core  platforms was  controlled using the  modules framework
which gives  an easy and  flexible mechanism to  access to all  of the
CSCS  provided compilers,  tools  and applications.   For our  initial
benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
a  good  level  of optimization  and  the  code  runs faster  in  this
configuration  than when  compiled with  the intel  compiler (14.0.1).
Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
libraries in favor of the  traditional GRIB library for the management
of extremely  large and complex data collections.   All computes nodes
have    an   operating   system    based   on    GNU/Linux   featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

>>>>

\subsection{Environment setup}
\label{subsec:4.2}
While hardware  platforms mature  and are replaced,  we have  chosen a
state-of-the-art  Intel's  third-generation   Core  (aka  Ivy  Bridge)
processing platform (called ``Monch'')  for our power measurements, as
it  is slated  to stay  in service  without hardware  upgrade  for the
duration  of  the Exa2Green  project.   In  principle  at least,  this
architecture could be recreated or found in an identical configuration
beyond the lifetime of the project.  Given that the baseline benchmark
can be reproduced  within an expected variance, and  that the baseline
run configuration  can be used in  all future versions of  the code, a
fair comparison  will be made  between the baseline and  the milestone
versions   of    COSMO-ART.    In   this    study,   a   complementary
energy-to-solution  benchmarking  comparison  is  carried out  on  the
``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
processors, conventional  in HPC  systems and known  to be  more power
consuming.   Finally a  full tracing  experiment is  conducted  on the
``Tintorrum''  cluster using  an integrated  framework, to  capture an
overall  power profile  at a  much finer  resolution and  to  get more
insights  into  the  MPI  blocking  and polling  influences  on  power
savings.

\subsubsection{Monch (CSCS - ETH Zurich)}
Monch, which is installed  at the Swiss National Supercomputing Center
(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
Ivy Bridge-EP based cluster, utilised from people that are part of the
Swiss    Platform   for    Advanced   Scientific    Computing   (PASC,
\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
base  clock speed,  themselves connected  by a  high  speed InfiniBand
network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
256 KB  of L2 cache. All  the 10 core share  a 25 MB L3  cache and the
platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
benchmark, a  full rack  of Monch constituted  of 52  standard compute
nodes (monchc[029-080]) was considered.

\subsubsection{Pilatus (CSCS - ETH Zurich)}
Pilatus,  which  is installed  at  the  Swiss National  Supercomputing
Center (CSCS) of  ETH Zurich, is a dual-socket  eight-core Intel Sandy
Bridge EP based cluster used as Piz Daint pre-post processing cluster.
It is composed of 42  compute nodes and has 2 high-speed interconnects
based on FDR: the first is dedicated to the MPI traffic and the second
to  the storage  high speed  traffic.  The  2 login  nodes and  the 42
computes nodes consists in 11 twin-pair Intel E5-Series DALCO r2264i4t
2U  scalable compute modules.   Each module  contains 4  compute nodes
based on two  Intel Xeon E5-2670 processors operating  at 2.6 GHz base
clock speed,  themselves connected by a high  speed InfiniBand network
based  on  Mellanox  SX6036  managed  FDR switches,  with  a  56  Gb/s
speed. Each  core has 32 KB instruction  and 32 KB data  L1 caches and
256 KB  of L2 cache.  All  the 8 core share  a 20 MB L3  cache and the
platform has  64GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
benchmark, a full  rack of Pilatus constituted of  42 standard compute
nodes (pilatus[03-44]) was considered.

@@

\subsubsection{Sofware environment}
INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
distributed  memory  parallel  computers  using  the  Message  Passing
Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
both  many-core  CSCS  platforms  was  controlled  using  the  modules
framework which gives an easy  and flexible mechanism to access to all
of the  provided compilers, tools  and applications.  For  our initial
benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
a  good  level  of optimization  and  the  code  runs faster  in  this
configuration  than when  compiled with  the intel  compiler (14.0.1).
Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
libraries in favor of the  traditional GRIB library for the management
of extremely  large and complex data collections.   All computes nodes
have    an   operating   system    based   on    GNU/Linux   featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

### 2014-04-18 paper_enahpc14/paper.tex
adds very few changes; adds Cristiano Malossi as a co-author

\author{Joseph~Charles  \and William~Sawyer  \and  Manuel~F.~Dolz \and
  Sandra~Catal\'an}

@@

\institute{J.~Charles, W.~Sawyer 
	\at Swiss National Supercomputing Centre (CSCS)
	\\ CH-6900 Lugano, Switzerland  
 	\\ E-mails:~{\{joseph.charles,william.sawyer\}@cscs.ch} 
        \and M.~F.~Dolz 
        \at Dept. of Informatics, University of Hamburg (UHAM)
	\\ DE-22527 Hamburg, Germany
        \\ \email{manuel.dolz@informatik.uni-hamburg.de}  
        \and S.~Catal\'an
        \at Jaume I University of Castell\'on (UJI)
	\\ ES-12071 Castell\'on, Spain
        \\ \email{catalans@icc.uji.es}}

@@

\begin{abstract}
  In this paper we present  COSMO-ART, an extension of the operational
  weather  forecast  model  of   the  German  weather  service  (DWD),
  developed for  the evaluation of the interactions  of reactive gases
  and aerosol particles  with the state of atmosphere  at the regional
  scale. It  includes secondary aerosols,  directly emitted components
  like  soot,  mineral  dust,  sea  salt and  biological  material  as
  pollen. Processes such  as emissions, coagulation, condensation, dry
  deposition,  wet removal,  and sedimentation  of aerosols  are taken
  into account.   The overall performance  of this application  on HPC
  systems is analysed  by a profiling study to  determine hotspots and
  identify  critical  CPU paths.   Moreover,  we describe  measurement
  devices and energy-aware techniques  employed to evaluate the energy
  footprint of the considered application and to get detailed insights
  about power bottlenecks.  Our motivation is to improve corresponding
  code   sections  to  sustain   high  performance   while  minimizing
  energy-to-solution.   This  preliminary  work  sets  the  basis  for
  subsequent studies to tackle  challenges related to energy efficient
  high performance computing in the framework of the Exa2Green project
  (\url{http://exa2green.eu/}).

\keywords{High performance computing  \and Energy-aware computing \and
  Green computing  \and Numerical weather  prediction \and Atmospheric
  chemistry  \and  Aerosols  modelling  \and  Profiling  methods  \and
  Benchmark    analysis   \and    COSMO-ART    coupled   model    \and
  Time-to-solution \and Energy-to-solution \and Manycore}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{sec:1}
\input{Sec1:Introduction.tex}

@@

\section{Experimental setup}
\label{sec:3}
\input{Sec3:Power-measurement.tex}

@@

\section{Related work}
\label{sec:5}
\input{Sec5:Related.tex}

\section{Conclusion}
\label{sec:6}
\label{Sec6:Conclusions.tex}
\input{Conclusion.tex}

>>>>

\author{Joseph~Charles \and William~Sawyer \and Manuel~F.~Dolz \and
  Sandra~Catal\'an \and Cristiano~Malossi }

@@

\institute{J.~Charles, W.~Sawyer 
	\at Swiss National Supercomputing Centre (CSCS)
	\\ CH-6900 Lugano, Switzerland  
 	\\ E-mails:~{\{joseph.charles,william.sawyer\}@cscs.ch} 
        \and M.~F.~Dolz 
        \at Dept. of Informatics, University of Hamburg (UHAM)
	\\ DE-22527 Hamburg, Germany
        \\ \email{manuel.dolz@informatik.uni-hamburg.de}  
        \and S.~Catal\'an
        \at Jaume I University of Castell\'on (UJI)
	\\ ES-12071 Castell\'on, Spain
        \\ \email{catalans@icc.uji.es}
        \and C.~Malossi  
        \at IBM Zurich Research Laboratory 
        \\ CH-8000 Zurich, Switzerland
        \\ \email{ACM@zurich.ibm.com}
}

@@

\begin{abstract}
  In  this paper  we present  the energy-to-solution  of  COSMO-ART on
  various platforms.   This model is  an extension of  the operational
  weather  forecast  model  of   the  German  weather  service  (DWD),
  developed for  the evaluation of the interactions  of reactive gases
  and aerosol particles  with the state of atmosphere  at the regional
  scale. It  includes secondary aerosols,  directly emitted components
  like  soot,  mineral  dust,  sea  salt and  biological  material  as
  pollen. Processes such  as emissions, coagulation, condensation, dry
  deposition,  wet removal,  and sedimentation  of aerosols  are taken
  into account.   The overall performance  of this application  on HPC
  systems is analysed  by a profiling study to  determine hotspots and
  identify  critical  CPU paths.   Moreover,  we describe  measurement
  devices and energy-aware techniques  employed to evaluate the energy
  footprint of the considered application and to get detailed insights
  about power bottlenecks.  Our motivation is to improve corresponding
  code    sections   to    sustain   performance    while   minimizing
  energy-to-solution.   This  preliminary  work  sets  the  basis  for
  subsequent studies to tackle  challenges related to energy efficient
  high performance computing in the framework of the Exa2Green project
  (\url{http://exa2green.eu/}).

\keywords{High performance computing  \and Energy-aware computing \and
  Green computing  \and Numerical weather  prediction \and Atmospheric
  chemistry  \and  Aerosols  modelling  \and  Profiling  methods  \and
  Benchmark    analysis   \and    COSMO-ART    coupled   model    \and
  Time-to-solution \and Energy-to-solution \and Xeon processor}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}
\input{Introduction.tex}

\section{Related work}
\label{sec:1}
\input{Sec1:Related.tex}

@@

\section{Power-performance measurement framework}
\label{sec:3}
\input{Sec3:Measurement-framework.tex}

@@

\section{Conclusion}
\label{concl}
\input{Conclusion.tex}
